High-order cognition is supported by information-rich but compressible brain activity patterns

### 2023-03-08 paper/main.tex
added stats, updated figures 2 and 3 (including captions)

  \includegraphics[width=0.8\textwidth]{figs/decoding_and_inflection}

\caption{\textbf{Decoding accuracy and compression.} \textbf{A. Decoding
accuracy by number of components.} Ribbons of each color display
cross-validated decoding performance for each condition (intact, paragraph,
word, and rest), as a function of the number of components (features) used to
represent the data. The horizontal red line denotes chance performance, and the
horizontal black line denotes 5\% decoding accuracy (used as a reference in
Panel B). \textbf{B. Numbers of components required to reach a fixed decoding
accuracy threshold, by condition.} The panel displays a zoomed-in view of the
inset in Panel A. Intersections between each condition's decoding accuracy
curve and the 5\% decoding accuracy reference line are marked by arrows.
\textbf{C. Estimating inflection points.} We sought to identify an ``inflection
point'' for each decoding curve, denoting the number of components at which the
decoding curve asymptotes. We fit sigmoid functions to each decoding curve
(left sub-panel) and then computed the minimum number of components for which
the second derivative of the sigmoid was both positive and less than a
threshold value of 0.0001. \textbf{D. Inflection points by condition.} Each dot
displays the number of components ($x$-axis) and decoding accuracy ($y$-axis)
at one condition's inflection point. All error ribbons denote
bootstrap-estimated 95\% confidence intervals.}

@@

To evaluate the relation between informativeness and compressibiilty for brain
activity from each experimental condition, we trained a series of
across-participant temporal decoders on compressed representations of the data.
Figure~\ref{fig:inflection}A displays the decoding accuracy as a function of
the number of principal components used to represent the data. Several patterns
were revealed by the analysis. First, in general (i.e., across experimental
conditions), decoding accuracy improves as the number of components increases.
However, decoding accuracy peaked at higher levels for experimental conditions
that exposed participants to cognitively richer stimuli. The peak decoding
accuracy was highest for the ``intact'' condition (versus paragraph: $t(XXX) =
XXX, p = XXX$; versus word: $t(XXX) = XXX, p = XXX$; versus rest: $t(XXX) =
XXX, p = XXX$), next highest for the ``paragraph'' condition (versus word:
$t(XXX) = XXX, p = XXX$; versus rest: $t(XXX) = XXX, p = XXX$), and next
highest for the ``word'' condition (versus rest: $t(XXX) = XXX, p = XXX$). This
ordering implies that cognitively richer conditions evoke more stable brain
activity patterns across people.

The cognitively richer conditions also displayed steeper initial slopes. For
example, the intact condition decoders reached an arbitrarily chosen threshold
of 5\% accuracy using fewer components than the paragraph condition decoders
($t(XXX) = XXX, p = XXX$) or word condition decoders ($t(XXX) = XXX, p = XXX$),
and decoding accuracy never exceeded 5\% for the rest condition. This suggests
that brain activity patterns evoked by cognitively richer conditions are more
compressible, such that representing the data using the same number of
principal components provides more information to the temporal decoders
(Fig.~\ref{fig:inflection}B).

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=0.8\textwidth]{figs/information_and_compression_over_time}

\caption{\textbf{Changes in decoding accuracy and compression over time.}
\textbf{A. Decoding accuracy by number of components, by story segment.} Each
family of curves is plotted in the same format as Figure~\ref{fig:inflection}A
but reflects data only from one third of the dataset. \textbf{B. Inflection
points by condition and segment.} The dots re-plot the inflection points from
Figure~\ref{fig:inflection}D for reference. The numbers denote the inflection
points for each third of the dataset (1: first third; 2: second third; 3: third
third; colors denote experimental conditions). \textbf{C. Change in decoding
accuracy over time, by number of components.} For each number of components
($x$-axis) and condition (color), we fit a regression line to the decoding
accuracies obtained for the first, second, and third thirds of the dataset
(corresponding to the left, middle, and right columns of Panel A,
respectively). The $y$-axis denotes the slopes of the regression lines. All
error ribbons denote bootstrap-estimated 95\% confidence intervals.}

@@

If informativeness (to the temporal decoders) and compressibility vary with the
cognitive richness of the stimulus, might these measures also vary over time
\textit{within} a given condition? For example, participants in the intact
condition might process the ongoing story more deeply later on in the story
(compared with earlier in the story) given the additional narrative background
and context they had been exposed to by that point. To examine this
possibility, we divided each condition into three successive time segments. We
computed decoding curves (Fig.~\ref{fig:inflection-thirds}A) and inflection
points (Fig.~\ref{fig:inflection-thirds}B) for each segment and condition. We
found that, in the two most cognitively rich conditions (intact and paragraph),
both decoding accuracy and compressibility, as reflected by the change in
decoding curves, increased with listening time (intact: $t(XXX) = XXX, p =
XXX$; paragraph: $t(XXX) = XXX, p = XXX$). These changes may reflect an
increase in comprehension or depth of processing with listening time. In
contrast, the decoding accuracy and compressibility \textit{decreased} with
listening time in the word condition ($t(XXX) = XXX, p = XXX$) and rest
condition ($t(XXX) = XXX, p = XXX$). This might reflect the depletion of
attentional resources in the less-engaging word and rest conditions.

>>>>

  \includegraphics[width=\textwidth]{figs/decoding_and_inflection}

\caption{\textbf{Decoding accuracy and compression.} \textbf{A. Decoding
accuracy by number of components.} Ribbons of each color display
cross-validated decoding performance for each condition (intact, paragraph,
word, and rest), as a function of the number of components (features) used to
represent the data. For each condition, decoding accuracy has been normalized
by subtracting expected ``chance'' decoding performance (estimated as
$\frac{1}{T}$, where $T$ is the number of timepoints in the given condition).
This normalization was intended to enable fairer comparisons across conditions;
a relative decoding accuracy of 0.0 denotes ``chance'' performance. The
horizontal black line denotes 5\% decoding accuracy (used as a reference in
Panel B). \textbf{B. Numbers of components required to reach a fixed decoding
accuracy threshold, by condition.} The panel displays a zoomed-in view of the
inset in Panel A. Intersections between each condition's decoding accuracy
curve and the 5\% decoding accuracy reference line are marked by arrows. All
error ribbons in Panels A and B denote bootstrap-estimated 95\% confidence
intervals. \textbf{C. Estimating ``compressibility'' for each condition.} The
density plots display the numbers of components required to reach at least 5\%
(corrected) decoding accuracy, across all cross validation runs. For the
``rest'' condition, where decoding accuracy never reached 5\% accuracy, we
display the distribution of the numbers of components required to achieve the
peak decoding accuracy in each fold. (This distribution is nearly flat, as
illustrated in Panel E.) \textbf{D. Estimating ``informativeness'' for each
condition.} The density plots display the peak (corrected) decoding accuracies
achieved in each condition, across all cross validation runs. \textbf{E.
Informativeness versus compressibility.} Each dot displays the minimum number
of components required to achieve at least 5\% corrected decoding accuracy
($x$-coordinate; for the rest condition the numbers of components are estimated
using the peak decoding accuracies as in Panel C) and the peak (corrected)
decoding accuracy ($y$-coordinate). The smaller dots correspond to individual
cross validation runs and the larger dots display the averages across runs. The
inset displays a zoomed-in view of the indicated region in the main panel.}

@@

To evaluate the relation between informativeness and compressibiilty for brain
activity from each experimental condition, we trained a series of
across-participant temporal decoders on compressed representations of the data.
Figure~\ref{fig:inflection}A displays the decoding accuracy as a function of
the number of principal components used to represent the data. Several patterns
were revealed by the analysis. First, in general (i.e., across experimental
conditions), decoding accuracy improves as the number of components increases.
However, decoding accuracy peaked at higher levels for experimental conditions
that exposed participants to cognitively richer stimuli. The peak decoding
accuracy was highest for the ``intact'' condition (versus paragraph: $t(99) =
35.205, p < 0.001)$; versus word: $t(99) = 43.172, p < 0.001)$; versus rest:
$t(99) = 81.361, p < 0.001)$), next highest for the ``paragraph'' condition
(versus word: $t(99) = 6.243, p < 0.001)$; versus rest: $t(99) = 50.748, p <
0.001)$), and next highest for the ``word'' condition (versus rest: $t(99) =
48.791, p < 0.001)$). This ordering implies that cognitively richer conditions
evoke more stable brain activity patterns across people.

The cognitively richer conditions also displayed steeper initial slopes. For
example, the intact condition decoders reached an arbitrarily chosen threshold
of 5\% accuracy using fewer components than the paragraph condition decoders
($t(99) = -7.429, p < 0.001)$) or word condition decoders ($t(99) = -7.300, p <
0.001)$), and decoding accuracy never exceeded 5\% for the rest condition. This
suggests that brain activity patterns evoked by cognitively richer conditions
are more compressible, such that representing the data using the same number of
principal components provides more information to the temporal decoders
(Fig.~\ref{fig:inflection}B).

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{figs/information_and_compression_over_time}

\caption{\textbf{Changes in decoding accuracy and compression over time.}
\textbf{A. Decoding accuracy by number of components, by story segment.} Each
family of curves is plotted in the same format as Figure~\ref{fig:inflection}A
but reflects data only from one quarter of the dataset. \textbf{B.
Informativeness versus compressibility by condition and segment.} Each scatter
plot is in the same format as Figure~\ref{fig:inflection}E, but reflects data
only from one quarter of the dataset. \textbf{C. Change in decoding accuracy
over time, by number of components.} For each number of components ($x$-axis)
and condition (color), we fit a regression line to the decoding accuracies
obtained for the first, second, third, and fourth quarters of the dataset
(corresponding to the columns of Panels A and B). The $y$-axis denotes the
slopes of the regression lines. All error ribbons denote bootstrap-estimated
95\% confidence intervals.}

@@

If informativeness (to the temporal decoders) and compressibility vary with the
cognitive richness of the stimulus, might these measures also vary over time
\textit{within} a given condition? For example, participants in the intact
condition might process the ongoing story more deeply later on in the story
(compared with earlier in the story) given the additional narrative background
and context they had been exposed to by that point. To examine this
possibility, we divided each condition into three successive time segments. We
computed decoding curves (Fig.~\ref{fig:inflection-thirds}A) and inflection
points (Fig.~\ref{fig:inflection-thirds}B) for each segment and condition. We
found that, in the two most cognitively rich conditions (intact and paragraph),
both decoding accuracy and compressibility, as reflected by the change in
decoding curves, increased with listening time (e.g., at the annotated
reference point of $k = 20$ components: intact: $t(99) = 7.915, p < 0.001)$;
paragraph: $t(99) = 2.354, p = 0.021)$). These changes may reflect an increase
in comprehension or depth of processing with listening time. In contrast, the
decoding accuracy and compressibility \textit{decreased} with listening time in
the word condition ($t(99) = -10.747, p < 0.001)$) and rest condition ($t(99) =
-22.081, p < 0.001)$). This might reflect the depletion of attentional
resources in the less-engaging word and rest conditions.
